Type theory in type theory with single substitutions

### 2024-09-07 types_2024_postproc/p.tex
stuff

\documentclass[a4paper]{article}

\usepackage{cite}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{hyperref}

@@

\title{Type theory in type theory with single substitutions}
\author{Ambrus Kaposi and Szumi Xie}
\date{\today}
\maketitle

\begin{abstract}
Type theory can be described as a generalised algebraic theory which
automatically gives a notion of model and the existence of the syntax
as the initial model, which is a quotient inductive-inductive
type. Such descriptions include Ehrhard's definition of model,
categories with families (CwF), contextual categories, Awodey's
natural models, C-systems and B-systems. With the exception of
B-systems, these notions are based on a parallel substitution calculus
where substitutions form a category. In this paper we define a single
substitution calculus (SSC) for type theory and show that the SSC
syntax and the CwF syntax are isomorphic for a theory with dependent
function space and universe. SSC only includes single substitutions
and single weakenings, and six equations relating these. Four
equations describe how to substitute variables, and two more equations
are needed to typecheck these four equations. SSC provides a simple,
minimalistic alternative to parallel substitution calculi or B-systems
for defining type theory. All the results in this paper were
formalised in Agda.
\end{abstract}

@@

\cite{singleTypes}
\cite{coquandEhrhard}

What is type theory? This term can refer to formal systems (languages,
type systems, type theories) or to the field of study of type
theories. In this paper, we will give an answer to this question in
the form of a concrete formal system. We aim for simplicity and ease
of understanding: we only require a working knowledge of Agda, a
functional programming language which is an implementation of type
theory. Our metatheory in which this paper is written is thus type
theory, and everything is formalised in Agda.

The traditional way of defining a programming language formally
\cite{DBLP:books/daglib/0005958,DBLP:books/cu/Ha2016,Pierce:SF1,plfa22.08}
is comprised of three steps: (i) definition of the presyntax
(abstract syntax trees, ASTs) as a formal grammar; (ii) specification of
well-formedness/well-typedness relations saying which ASTs are
well-behaved; (iii) specification of rewriting/conversion relations
explaining how programs are executed. Instead of following this route,
we use the intrinsic approach
\cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16}: (i)
we define ASTs as an inductive type instead of production rules of a
grammar; (ii) terms are indexed by (object theoretic) types, thus only
allowing ASTs to be plugged together in well-typed ways; (iii) terms
are quotiented by conversion rules, making convertible terms
equal. Inductive types which allow such mutual indexing and
quotienting are called quotient inductive-inductive types (QIITs
\cite{DBLP:journals/pacmpl/KaposiKA19}). As Agda does not support them
directly, we postulate their constructors and induction principle and add the
computation rules using rewrite rules \cite{DBLP:conf/types/Cockx19}.

A QIIT is also the initial algebra/model of a generalised algebraic
theory (GAT \cite{DBLP:journals/apal/Cartmell86}). The induction principle of
the QIIT is equivalent to having a unique morphism into any model
\cite{DBLP:journals/pacmpl/KaposiKA19}. There are several ways to
define type theory as a GAT: categories with families (CwFs
\cite{DBLP:journals/corr/abs-1904-00827}), contextual categories
\cite{DBLP:journals/apal/Cartmell86}, natural models
\cite{DBLP:journals/mscs/Awodey18}, C-systems and B-systems
\cite{ahrens_emmenegger_north_rijke_2023}. When extracting a QIIT from
a GAT, with the exception of B-systems, we obtain parallel
substitution calculi. Like B-systems, in this paper we define type
theory using \emph{single substitutions}. Our definition is simpler
than B-systems, but it is closely related to it. It is a minimalistic
presentation where every operation and equation is well-motivated (see
Section \ref{sec:SSC}). However, when we want to use our syntax (for
example, if we want to prove canonicity or normalisation for our
theory), it is much more convenient to use parallel substitution
calculi such as CwFs. We first show how to get rid of explicit
substitutions from the syntax using the technique of
$\alpha$-normalisation (Section \ref{sec:alpha}). Then we show how to
derive the rules of a CwF from our syntax (Section
\ref{sec:cwf}). Finally, we show that our syntax is isomorphic to the
CwF-style syntax (Section \ref{sec:iso}). We list related work and
conclude in Section \ref{sec:conclusion}.

@@

@@

@@

\section{Related work and conclusion}

>>>>

\documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}

\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\ext}{\mathop{\triangleright}}

@@

\title{Type theory in type theory with single substitutions}
\author{Ambrus Kaposi and Szumi Xie}
\date{\today}

\begin{abstract}
Type theory can be described as a generalised algebraic theory. This
automatically gives a notion of model and the existence of the syntax
as the initial model, which is a quotient inductive-inductive
type. Algebraic definitions of type theory include Ehrhard's
definition of model, categories with families (CwFs), contextual
categories, Awodey's natural models, C-systems, B-systems. With the
exception of B-systems, these notions are based on a parallel
substitution calculus where substitutions form a category. In this
paper we define a single substitution calculus (SSC) for type theory
and show that the SSC syntax and the CwF syntax are isomorphic for a
theory with dependent function space and a hierarchy of universes. SSC
only includes single substitutions and single weakenings, and eight
equations relating these: four equations describe how to substitute
variables, and there are four equations on types which are needed to
typecheck other equations. All the other equations are substitution
(naturality) rules or computation rules for different type
formers. SSC provides a simple, minimalistic alternative to parallel
substitution calculi or B-systems for defining type theory. SSC
relates to CwF as extensional combinatory calculus relates to lambda
calculus. All the results in this paper were formalised in Agda.
\end{abstract}

\maketitle

@@

What is type theory? Here we refer to type theory as a particular
formal system based on Martin-Löf's original definition
\cite{martinlof73predicative}, and not to the field studying type
systems (e.g.\ \cite{DBLP:books/daglib/0005958}). When
answering this question, we aim for simplicity and ease of
understanding. We only require a working knowledge of an
implementation of type theory such as Agda. Our metatheory in which
this paper is written is thus type theory, and everything is
formalised in Agda.

The traditional way of defining a programming language formally
\cite{DBLP:books/cu/Ha2016,Pierce:SF2,plfa22.08} is comprised of three
steps: (i) definition of the presyntax (abstract syntax trees, ASTs)
as a formal grammar; (ii) specification of
well-formedness/well-typedness relations saying which ASTs are
well-behaved; (iii) specification of rewriting/conversion relations
explaining how programs are executed. Instead of following this route,
we use the intrinsic approach
\cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16}: (i)
we define ASTs as an inductive type instead of production rules of a
grammar; (ii) terms are indexed by (object theoretic) types, thus only
allowing ASTs to be built in well-typed ways; (iii) terms are
quotiented by conversion rules, making convertible terms equal. The
advantage of the intrinsic approach that it provides shorter, easier
to understand, less technical definitions, and there is no separation
between syntax and semantics. Type theory is a generalised algebraic
theory (GAT \cite{DBLP:journals/apal/Cartmell86}), the notion of model
is direct, and the syntax is simply the initial model. GATs are
generalisations of single-sorted algebraic theories (such as monoids,
groups) to multiple sorts where later sorts can be indexed over
previous sorts. A typical example is categories with a sorts of
objects and a sort of morphisms double-indexed over objects. The
initial model of a GAT is a quotient inductive-inductive type (QIIT
\cite{DBLP:journals/pacmpl/KaposiKA19}). QIITs are a general family of
inductive types where multiple types are defined mutually, later types
can be indexed over previous ones, and equality constructors are also
allowed, generalising quotients. The elimination principle of the QIIT
corresponds to a structural induction principle for the QIIT and is
equivalent to the statement that the syntax is the initial model
\cite{DBLP:journals/pacmpl/KaposiKA19}, that is, there is a unique
morphism from the syntax to any other model.

There are several ways to define type theory as a GAT. For example,
Thomas Ehrhard's calculus \cite{ehrhard,coquandEhrhard}, Dybjer's
categories with families (CwFs
\cite{DBLP:conf/types/Dybjer95,Castellan2021}), contextual categories
or C-systems
\cite{DBLP:journals/apal/Cartmell86,DBLP:journals/lmcs/AhrensLV18},
natural models \cite{DBLP:journals/mscs/Awodey18}, B-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}. These are providing the
basic substitution calculus part of the theory, and individual type
formers (such as function space or product types) have to be specified
separately. With the exception of B-systems, the listed definitions
are all based on parallel substitution calculi, where contexts and
substitutions form a category. Like B-systems, in this paper we define
type theory using \emph{single substitutions}. Single substitutions
don't compose, so they do not form a category. Our definition is
simpler than B-systems, but it is closely related to it. It is a
minimalistic presentation where every operation and equation is
well-motivated (see Section \ref{sec:ssc}). In addition to its
pedagogical value due to its simplicity, our definition might
contribute in the future to a possible coherent syntax for type
theory, see Section \ref{sec:conclusion}.

Our theory is in fact a bit too minimalistic to be workable: for
example, consider a term $b$ which depends on a context $\Gamma$ and
an extra variable of type $A$, that is, $b : \Tm\,(\Gamma\ext
A)\,B$. We can weaken $b$ so that we add an extra $A$ variable
\emph{before} the last one obtaining $b[\p^+] : Tm\,(\Gamma\ext A\ext
A[\p])\,(B[\p])$ where $\p$ is the weakening substitution. Then, we
instantiate the last $A[\p]$ variable in the context by the previous
$A$ obtaining $b[\p^+][\langle\q\rangle]$ where $\q$ is De Bruijn
index $0$. We cannot derive $b[\p^+][\langle\q\rangle] = b$ in an
arbitrary model, but this equation is admissible in the \emph{syntax}.
% p⁺  : Sub (Γ▹A▹A[p]) (Γ▹A)
% ⟨q⟩ : Sub (Γ▹A) (Γ▹A▹A[p])
This is analogous to e.g.\ parametricity results
\cite{DBLP:journals/jfp/BernardyJP12}, which do not hold in an
arbitrary model, but they hold for syntactic terms. In fact, all the
operations and equations of the parallel substitution calculus (CwF)
are admissible in our syntax, moreover the set of single substitution
based syntactic terms are isomorphic to the set of CwF-based
terms. Every CwF-based model is also a model of our single
substitution calculus, but not the other way: there are more
SSC-models than CwF-models. The situation is analogous to that of
extensional combinatory calculus and lambda calculus
\cite{DBLP:conf/fscd/AltenkirchKSV23}: the syntactic terms form
isomorphic sets, but there are more combinatory models than lambda
calculus models.

TODO: rewrite from here. Add: discussion of type formers, SOGATs and say that we only do this
for an example t.t. and not for all of them. We conjecture that for
any SOGAT, the SSC and CwF-based syntaxes are equivalent.

The summary of the above discussion: we define a new minimal syntax
for type theory which is equivalent to the original syntax, but

We first show how to get rid of explicit substitutions from the syntax
using the technique of $\alpha$-normalisation (Section
\ref{sec:alpha}). Then we show how to derive the rules of a CwF from
our syntax (Section \ref{sec:cwf}). Finally, we show that our syntax
is isomorphic to the CwF-style syntax (Section \ref{sec:iso}). We list
related work and conclude in Section \ref{sec:conclusion}.

@@

% TODO: say that this is the paper version of the \cite{singleTypes}
% paper

@@

As Agda does not support them directly, we postulate their
constructors and induction principle and add the computation rules
using rewrite rules \cite{DBLP:conf/types/Cockx19}.

@@

\section{Conclusions}
\label{sec:conclusion}
